---
title: "EFISUR a new acquisition function (infill sampling criterion)"
author: "Reda El Amri"
date: "2021-03-14"
output: github_document
  #pdf_document: default
  #html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

We consider the problem of chance constrained optimization where it is sought to optimize a function and satisfy constraints, both of which are affected by uncertainties. The real world declinations of this
problem are particularly challenging because of their inherent computational cost.

We propose a new Bayesian optimization method. It applies to the situation where the uncertainty comes from
some of the inputs, so that it becomes possible to define an acquisition criterion in the joint controlled-uncontrolled input space (see the paper for more details <https://arxiv.org/pdf/2103.05706.pdf>).

## Load R-packages

```{r echo=FALSE}
rm(list=ls())

#library('DiceDesign', quietly = TRUE)
#library('DiceKriging', quietly = TRUE)
#library("nloptr", quietly = TRUE)
#library("randtoolbox", quietly = TRUE)

###################################################################################### 
fct1 <- function(input) return((1/3 * input[2]^4 - 2.1*input[2]^2 + 4)*input[2]^2 + 
                                 input[1]*input[2] + 4*input[1]^2*(input[1]^2-1))


fct2 <- function(input) return(((3*input[1]^2+7*input[1]*input[2]-3)*exp(-1*(input[1]*input[2])^2)*
                                  cos(5*pi*input[1]^2)-1.2*input[2]^2))


fct3 <- function(input) return( 5*(input[1]^2 + input[2]^2) - (input[3]^2 + input[4]^2) +
                                  input[1]*(-input[3] + input[4]+5) + input[2]*(input[3] - input[4] + 3) )


fct4 <- function(input) return( -input[1]^2 + 5*input[2] - input[3] + input[4]^2 - 1)

Objective_function <- function(x,d,m,names,model_joint_objective){
  control <- matrix(x, nrow = 1, ncol = (d+m))
  control <- data.frame(control)
  names(control) <- names[1:(d+m)]
  p <- predict(model_joint_objective, newdata = data.frame(x = control), type = "UK",checkNames=FALSE)
  return(p$mean)
}

Expectation_Objective_function <- function(x,alea,d,m,model_joint_objective){
  m1 <- matrix(x, nrow = 1, ncol = d)
  m1 =  t(matrix(rep(m1,dim(alea)[1]),d,dim(alea)[1]))
  dat <- data.frame(cbind(m1,alea))
  p <- predict(model_joint_objective, newdata = data.frame(x = dat), type = "UK",checkNames=FALSE)
  return(mean(p$mean))
}

Constraint_function <- function(x,d,m,names,model_joint_constraint){
  control <- matrix(x, nrow = 1, ncol = (d+m))
  control <- data.frame(control)
  names(control) <- names[1:(d+m)]
  p <- predict(model_joint_constraint, newdata = data.frame(x = control), type = "UK",checkNames=FALSE)
  return(p$mean)
} 

Quantile_Constraint_function <- function(x,alea,d,m,model_joint_constraint){
  m1 <- matrix(x, nrow = 1, ncol = d)
  m1 =  t(matrix(rep(m1,dim(alea)[1]),d,dim(alea)[1]))
  dat <- data.frame(cbind(m1,alea))
  p <- predict(model_joint_constraint, newdata = data.frame(x = dat), type = "UK",checkNames=FALSE)
  return(quantile(p$mean,prob=0.95))
}

mean_sd_Z <- function(x,model,alea,d,m,select){
  x <- matrix(x,ncol=d,nrow=1)
  alea <- matrix(alea,ncol=m)
  m1 <- matrix(x, nrow = 1, ncol = d)
  m2 = t(matrix(rep(x,dim(alea)[1]),d,dim(alea)[1]))
  dat <- data.frame(cbind(m2,alea))
  pred <- DiceKriging::predict.km(model,dat,checkNames = FALSE,type="UK",cov.compute=FALSE)
  if(select=="TRUE") return(mean(pred$mean))
  if(select=="FALSE") return(abs(mean(pred$sd)))
}

Phi_G <- function(x,modelConst,alea,seuil,d,m,sign){
  input <- matrix(0,dim(alea)[1],d+m)
  input[,((d+1):(d+m))] <- alea
  for(j in 1:dim(alea)[1]) input[j,1:d] <- x
  pred <- DiceKriging::predict.km(modelConst,input,checkNames = FALSE,type="UK",cov.compute=FALSE)
  if(sign == ">") phi <- pnorm((pred$mean-seuil)/pred$sd)
  else phi <- pnorm((seuil-pred$mean)/pred$sd)
  return(phi)
}

Expectation_A <- function(x,alea,d,m,list_const_km,seuil,sign,alpha){
  d_c <- length(list_const_km) # constraints models
  PHI <- matrix(NA,nrow=d_c,ncol=dim(alea)[1])
  for(i in 1:d_c) PHI[i,] <- Phi_G(x,list_const_km[[i]],alea,seuil[i],d,m,sign[i])
  E_A <- 1 - alpha - mean(apply(PHI,2,prod))
  return(E_A)
}

z_mean_feasible <- function(modelObjective,d,m,alea,list_const_km,seuil,sign,alpha){
  X <- matrix(modelObjective@X[,1:d],ncol=d)
  res <- apply(X,1,mean_sd_Z,modelObjective,alea,d,m,select="TRUE")
  res1 <- apply(X,1,Expectation_A,alea,d,m,list_const_km,seuil,sign,alpha)
  index <- which(res1 <= 0)
  if(length(index) > 0 ) {index2 <- which.min(res[index]);   return(list(min=res[index[index2]],opt=index[index2],mar=1))}
  if(length(index) == 0 ) {t <- which.max(res1); return(list(min=res[t],opt=t,mar=2))}
}

Probability_A <- function(x,alea,Nsim,d,m,list_const_km,seuil,sign,alpha){
  input <- matrix(0,dim(alea)[1],d+m)
  input[,(d+1):(d+m)] <- alea
  for(j in 1:dim(alea)[1]) input[j,1:d] <- x
  d_c <- length(list_const_km) # constraints models
  realizations <- array(0, dim=c(Nsim,dim(alea)[1],d_c) ) # multidimensional array (Nsim,CRN,nbreConstraint)
  for(i in 1:d_c) realizations[,,i] <- DiceKriging::simulate(list_const_km[[i]],nsim=Nsim,newdata=input,
                                                             checkNames = FALSE,type="UK",cond=TRUE)
  for(i in 1:d_c)
  {
    if(sign[i] == ">") realizations[,,i] <-  seuil[i] - realizations[,,i]
    else realizations[,,i] <- realizations[,,i] - seuil[i]
  }
  sim <- apply(realizations,c(1,2),max) ## max(g1,g2,g3)
  p_sim <- apply(sim,1,function(vect) length(which(vect<=0))/dim(alea)[1])
  c <- length(which(1 - alpha - p_sim <= 0))/Nsim
  return(c=c)
}

Feas_Expected_Improvement <- function(x,minimum_feas,model,alea,Nsim,d,m,list_const_km,seuil,sign,alpha){
  sdZ <- abs(mean_sd_Z(x,model,alea,d,m,select="FALSE"))
  espZ <- mean_sd_Z(x,model,alea,d,m,select="TRUE")
  v <- (minimum_feas-espZ)/sdZ
  phi <- dnorm(v, mean = 0, sd = 1, log = FALSE)
  PHI <- pnorm(v, mean = 0, sd = 1, lower.tail = TRUE, log.p = FALSE)  
  EIZ <- sdZ*(v*PHI+phi)
  proba <- Probability_A(x,alea,Nsim,d,m,list_const_km,seuil,sign,alpha)
  return(c(EIZ)*proba)
}

PredVar_Z <- function(xstar,xp1,model,alea,d,m){
  U_MC <- dim(alea)[1]
  xxstar <- xstar
  objectinit <- model@covariance
  
  X1 <- (as.matrix( rbind(model@X,as.vector(xp1))))
  X2 <- (as.matrix( rbind(model@X,as.vector(xp1))))
  n1 <- nrow(X1)
  n2 <- nrow(X2)
  dim <- ncol(X1)
  C <- covMat1Mat2(model@covariance,X1,X2,nugget.flag = TRUE)/model@covariance@sd2 
  C <- solve(C)
  
  ########################################################### FIRST TERM
  U1 <- alea
  U2 <- alea
  nU1 <- nrow(U1)
  nU2 <- nrow(U2)
  dimU <- ncol(U1)
  object <- objectinit
  object@range.val <- covparam2vect(objectinit)[(d+1):(d+m)]
  outU <- covMat1Mat2(object,U1,U2,nugget.flag = TRUE)/model@covariance@sd2 
  MU <- mean(matrix(outU, nU1, nU2))
  term1 <- MU
  
  ########################################################### SECOND TERM
  
  X1 <- cbind(t(matrix(rep(xstar,U_MC),length(xstar),U_MC)),alea)
  X2 <- (as.matrix( rbind(model@X,as.vector(xp1))))
  n1 <- nrow(X1)
  n2 <- nrow(X2)
  dim <- ncol(X1)
  object <- objectinit
  object@range.val <- covparam2vect(objectinit)
  M <- covMat1Mat2(model@covariance,X1,X2,nugget.flag = TRUE)/model@covariance@sd2
  
  X1 <- (as.matrix( rbind(model@X,as.vector(xp1))))
  X2 <- cbind(t(matrix(rep(xstar,U_MC),length(xstar),U_MC)),alea) 
  n1 <- nrow(X1)
  n2 <- nrow(X2)
  dim <- ncol(X1)
  object <- objectinit
  object@range.val <- covparam2vect(objectinit)
  trans_M <- covMat1Mat2(model@covariance,X1,X2,nugget.flag = TRUE)/model@covariance@sd2
  
  a <- matrix(apply(M,2,mean),nrow=1)
  b <- matrix(apply(trans_M,1,mean),ncol=1)
  term2 <- a%*%C%*%(b)
  ###########################################################   
  return(term1 - term2)
}  

PredMean_Z <- function(x,xp1,observation,model,alea,d,m){
  m1 <- matrix(x, nrow = 1, ncol = d)
  m2 = t(matrix(rep(x,dim(alea)[1]),d,dim(alea)[1]))
  dat <- data.frame(cbind(m2,alea))
  colnames(dat) <- NULL
  xp1 <- as.vector(xp1)
  data <- data.frame(rbind(xp1,dat))
  utile <- DiceKriging::predict.km(model,data,checkNames = FALSE,type="UK",cov.compute=TRUE)
  term1 <- mean_sd_Z(x,model,alea,d,m,"TRUE")
  term2 <- (observation - utile$mean[1])/utile$sd[1]^2
  term3 <- mean(utile$cov[1,])  
  return(term1 + term2*term3)
} 

PredVar_F <- function(newdata,xp1,model,d,m){
  newdata <- matrix(newdata,ncol=d+m)
  xp1 <- matrix(xp1,ncol=d+m)
  utile <- DiceKriging::predict.km(model,rbind(xp1,newdata),checkNames = FALSE,type="UK",cov.compute=TRUE)
  return(utile$sd[2]^2 - utile$cov[1,2]^2/utile$sd[1]^2)
}

TotalVar <- function(x,xp1,model,alea,d,m,observation,ming,variZplus1){
  meanZplus1 <- PredMean_Z(x=x,xp1=xp1,observation,model=model,alea=alea,d=d,m=m)
  
  term <- (ming - meanZplus1)/sqrt(variZplus1)
  a <- pnorm(term, mean = 0, sd = 1)
  b <- dnorm(term, mean = 0, sd = 1)
  
  term1 <- (ming - meanZplus1)^2 + variZplus1
  term2 <- sqrt(variZplus1)*(ming - meanZplus1)
  
  expected_imprmnt <- (ming - meanZplus1)*a + sqrt(variZplus1)*b
  
  return(c(expected_imprmnt,term1*a + term2*b - expected_imprmnt^2))
}

aggregate_u_tp1 <- function(utilde,xnew,inputsss,rep,ming,modelObjective,alea,m,d,list_const_km,seuil,sign){
  input4s <- c(xnew,utilde)
  stopping <- 0
  oldlaw <- DiceKriging::predict.km(modelObjective,rbind(input4s),checkNames = FALSE,type="UK",cov.compute=TRUE)
  oldlaw <- c(oldlaw$mean,oldlaw$sd)
  aleaa <- matrix(randtoolbox::sobol(n=rep, dim = 1, init = TRUE, scrambling = 0, seed = 4711, normal = TRUE),ncol=1)
  aleaa <- aleaa*oldlaw[2] + oldlaw[1]
  variZplus1 <- abs(PredVar_Z(xnew,input4s,modelObjective,alea,d,m))
  term1 <- term2 <- NULL
  repeat{
    stopping = stopping+1
    muet <- TotalVar(xnew,input4s,modelObjective,alea,d,m,aleaa[stopping],ming,variZplus1)
    term1 <- c(term1,muet[1]) ;   term2 <- c(term2,muet[2])
    if (stopping == rep) break
  }
  res <- var(term1) + mean(term2)
  return(res)
  
  d_c <- length(list_const_km) # constraints models
  PHI <- matrix(NA,nrow=d_c,ncol=dim(inputsss)[1])
  for(i in 1:d_c)
  {
    pred <- DiceKriging::predict(list_const_km[[i]],newdata=inputsss,checkNames = FALSE,type="UK",cond=TRUE)
    predicted_sd2 <- apply(inputsss,1,PredVar_F,input4s,list_const_km[[i]],d,m)
    if(sign[i] == ">") PHI[i,] <- pnorm((pred$mean-seuil[i])/sqrt(abs(predicted_sd2)))
    else PHI[i,] <- pnorm((seuil[i]-pred$mean)/sqrt(abs(predicted_sd2)))
  }
  pq <- apply(PHI,2,prod)
  pp <- mean(pq*(1-pq))
  return(pp*res)
}

EFISUR <- function(model_O,listCon,bornes,d,m,alea,N,alpha,sign,seuil,solver){
  
  min_feas_init <- z_mean_feasible(model_O,d,m,alea,listCon,seuil,sign,alpha) # feasible minimum (z_min^{feas})
  hist_feasmin <- c(min_feas_init$min)  # feasible minimum value 
  hist_feasopt <- c(min_feas_init$opt) # associated point 
  
  f <- function(x) return(-Feas_Expected_Improvement(x,min_feas_init$min,model_O,alea,N,d,m,listCon,seuil,sign,alpha))
  
  if(solver == "bobyqa"){
    start = lhsDesign(1,d)$design
    for(a in 1:d) start[,a] <- bornes[a,1] + (bornes[a,2] - bornes[a,1])*start[,a]
    
    opt_solver_X <- bobyqa(start, fn = f,lower = bornes[1:d,1], upper = bornes[1:d,2],
                           control=list(maxeval = 500))
    x_tplus1 <- opt_solver_X$par
  }
  if(solver == "discrete"){
    N_xOpt <- 100*d
    Opt = lhsDesign(N_xOpt,d)$design
    for(a in 1:d) Opt[,a] <- bornes[a,1] + (bornes[a,2] - bornes[a,1])*Opt[,a]
    crit <- apply(Opt,1,f)
    x_tplus1 <- Opt[which.max(crit),]
  }
  
  ##### bobyqa in U
  
  inputsss <- matrix(0,N,(d+m))
  inputsss[,(d+1):(d+m)] <- alea
  for(j in 1:N) inputsss[j,1:d] <- x_tplus1
  
  g <- function(utilde) return(aggregate_u_tp1(utilde,x_tplus1,inputsss,rep=5,min_feas_init$min,model_O,
                                               alea,m,d,listCon,seuil,sign)) 
  
  if(solver == "bobyqa"){
    start = lhsDesign(1,m)$design
    for(a in 1:m) start[,a] <- bornes[(d+a),1] + (bornes[(d+a),2] - bornes[(d+a),1])*start[,a]
    opt_solver_U <- bobyqa(start, fn = g,lower = rep(bornes[(d+1):(d+m),1],m), 
                           upper = rep(bornes[(d+1):(d+m),2],m),control=list(maxeval = 200))
    u_tplus1 <- opt_solver_U$par
  }
  if(solver == "discrete"){
    N_uOpt <- 100*d
    Opt = lhsDesign(N_uOpt,d)$design
    for(a in 1:m) Opt[,a] <- bornes[(a+d),1] + (bornes[(a+d),2] - bornes[(a+d),1])*Opt[,a]
    crit <- apply(Opt,1,g)
    u_tplus1 <- Opt[which.min(crit),]
  }
  return(list(res=cbind(x_tplus1,u_tplus1),
              FeasMin=min_feas_init$min,
              OptPoint=model_O@X[min_feas_init$opt,1:d]))
}

```

```{r echo=TRUE}
library('DiceDesign', quietly = TRUE)
library('DiceKriging', quietly = TRUE)
library("nloptr", quietly = TRUE)
library("randtoolbox", quietly = TRUE)
```

## Initialization 

```{r echo=TRUE}
Objective_function <- fct1
Constraint_function <- list(fct2) # as a list

d <- 1 # dim of the control space
m <- 1 # dim of the uncertain space
bornesControl <- rbind(c(-1,1)) #X-space
bornesUncert <- rbind(c(0.35,0.65)) # U-space
bornes <- rbind(bornesControl,bornesUncert) #joint-space
seuil <- -2 # the thresholds (a vector if multiple constraints, e.g., c(3,4,6,9))
sign <- "<" # a vector if multiple constraints, e.g., c(">",">",">","<","<")
alpha <- 0.05 # parameter that controls the probability of constraint satisfaction
N <- 200 # realizations to estimate the probability of A
M_CRN <- 50 # CRN (Commmun Random Numbers)
alea <- matrix(randtoolbox::sobol(M_CRN,m),ncol=m)
for(i in 1:m) alea[,i] <- bornes[(i+d),1] + (bornes[(i+d),2] - bornes[(i+d),1])*alea[,i]

########################################################
##############  Initial Design of Experiments (DoE)
########################################################

n = 30*(d+m) # initial design size
DoE = lhsDesign(n,d+m)$design # initial LHS design in the joint space
DoE  <- maximinSA_LHS(DoE)$design # maximin-LHS design in the joint space
for(i in 1:(d+m)) DoE[,i] <- bornes[i,1] + (bornes[i,2] - bornes[i,1])*DoE[,i]

########################################################
##############  Initial Gaussian Process models
########################################################
 
d_c <- length(Constraint_function) # number of constraint functions
listCon <- NULL
response_f <- apply(DoE, 1, Objective_function) # objective function responses f(x,u)
model_O <- km(formula= ~1, design = data.frame(x=DoE), response= response_f,estim.method="LOO",
              control = list(trace=F),lower=rep(1e-10,d+m)) # GP of the objective function
for(l in 1:d_c){
  response_const <- apply(DoE, 1, Constraint_function[[l]]) # constraint function responses g(x,u)
  model_Const <- km(formula= ~1, design = data.frame(x=DoE), response= response_const,estim.method="LOO",
                    control = list(trace=F),lower=rep(1e-10,d+m)) # GP of the constraint functions
  listCon[[l]] <- model_Const
}

```

## Call the acquisition function: EFISUR 

```{r echo=TRUE}
solver <- "bobyqa" # choose the solver ("bobyqa or "discrete")
result <- EFISUR(model_O,listCon,bornes,d,m,alea,N,alpha,sign,seuil,solver)
```

## Plot

```{r echo=FALSE}
MC_U <- 200  # 2D example
aleatrue <- matrix(randtoolbox::sobol(MC_U,m),ncol=m)
for(i in 1:m) aleatrue[,i] <- bornesUncert[i,1] + (bornesUncert[i,2] - bornesUncert[i,1])*aleatrue[,i]
design.grid <- cbind(seq(-1,1,length=500))
true_Ef <- true_Pg <- rep(NA,dim(design.grid)[1])
for(i in 1:dim(design.grid)[1])
{
  input <- matrix(0,MC_U,d+m)
  input[,((d+1):(d+m))] <- aleatrue
  df <- data.frame(design.grid[i,]) 
  input[,1:d] <- as.matrix(df[rep(seq_len(nrow(df)), each=MC_U),],ncol=d)
  pred <- apply(input,1,Constraint_function[[1]])
  true_Ef[i] <- mean(apply(input,1,Objective_function))
  true_Pg[i] <- quantile(pred,prob=0.95)
}

id <- which(true_Pg<=seuil)
id1 <- which.min(true_Ef[id])
cost <- true_Ef[id[id1]]
opt.design <- design.grid[id[id1],]

plot(design.grid,true_Pg,type='l',xlab="x",ylab="Outputs")
lines(design.grid,true_Ef,col='red')
text(0,1,labels="red curve: E[f(x,U)]",pos=3,col='red',cex=0.7)
text(0.1,-3,labels="black curve: quantile[g(x,U),1-alpha]",pos=4,col='black',cex=0.7)
text(0.6,-1.8,labels="Threshold",pos=4,col='black')

abline(h=seuil,lwd=3,lty=2)
points(result$OptPoint,seuil,pch=19,col='blue')
text(result$OptPoint,seuil,labels="Next point",pos=3,col='blue')

print(paste(c("Next point (x,u) = ",as.vector(result$res)), collapse=" "))
print(paste("Best point x = ",result$OptPoint))
print(paste("cost = ",result$FeasMin))
```
